---
title: "2022 Standard Deviation of Redox Data 2022"
author: "Erin C Rooney"
date: "2023-06-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

load packages
```{r}
source("code/0-packages.R")
source("code/functions-for-std-dev-redox-data.R")

```


load metadata and data from each site. 
eventually we will combine everything, but we're keeping it separate for now.
```{r}

westhydric_dlname = read.csv("processed/westhydric_dlname2022.csv")
easthydric_dlname = read.csv("processed/easthydric_dlname2022.csv")
westmesic_dlname = read.csv("processed/westmesic_dlname2022.csv")
eastmesic_dlname = read.csv("processed/eastmesic_dlname2022.csv")
westdry_dlname = read.csv("processed/westdry_dlname2022.csv")
eastdry_dlname = read.csv("processed/eastdry_dlname2022.csv")

westhydric_metadata = read.csv("processed/westhydric_metadata2022.csv")
easthydric_metadata = read.csv("processed/easthydric_metadata2022.csv")
westmesic_metadata = read.csv("processed/westmesic_metadata2022.csv")
eastmesic_metadata = read.csv("processed/eastmesic_metadata2022.csv")
westdry_metadata = read.csv("processed/westdry_metadata2022.csv")
eastdry_metadata = read.csv("processed/eastdry_metadata2022.csv")

```

run the function on all 6 dataframes (for each site)
after running the function, we will pivot_longer so that the data is in two columns
data label including depth and probe
then we will "fix" the redox values by adding 197
then we will filter the data so that only fixed average values between -600 and 1000 are kept
```{r}

westhydric_avg = 
  avg_separate(westhydric_dlname) %>% 
  left_join(westhydric_metadata, by = 'redox_NUM_Avg') %>% 
  pivot_longer(-c("redox_NUM_Avg", "X", "redox_NUM_Std", "TIMESTAMP", "RECORD", "site", "position", "Betterdate"),
               names_to = "redox_avg", values_to = "avg_values") %>% 
  #add 197 to all redox potentials to report data relative to the standard hydrogen electrode
  filter(avg_values != 'NAN') %>% 
  mutate(avg_values = as.numeric(avg_values)) %>% 
  dplyr::mutate(avg_values_fixed = avg_values + 197) %>% 
  mutate(datalogger = "6")%>% 
  separate(redox_avg, sep = "_Av", into = c("redox_depth", "avg_delete")) %>% 
  dplyr::select(-avg_delete)

westmesic_avg = 
  avg_separate(westmesic_dlname) %>% 
  left_join(westmesic_metadata, by = 'redox_NUM_Avg') %>% 
  pivot_longer(-c("redox_NUM_Avg", "X", "redox_NUM_Std", "TIMESTAMP", "RECORD", "site", "position", "Betterdate"),
               names_to = "redox_avg", values_to = "avg_values") %>% 
  #add 197 to all redox potentials to report data relative to the standard hydrogen electrode
  filter(avg_values != 'NAN') %>% 
  mutate(avg_values = as.numeric(avg_values)) %>% 
  dplyr::mutate(avg_values_fixed = avg_values + 197) %>% 
  mutate(datalogger = "2")%>% 
  separate(redox_avg, sep = "_Av", into = c("redox_depth", "avg_delete")) %>% 
  dplyr::select(-avg_delete)

westdry_avg = 
  avg_separate(westdry_dlname) %>% 
  left_join(westdry_metadata, by = 'redox_NUM_Avg') %>% 
  pivot_longer(-c("redox_NUM_Avg", "X", "redox_NUM_Std", "TIMESTAMP", "RECORD", "site", "position", "Betterdate"),
               names_to = "redox_avg", values_to = "avg_values") %>% 
  #add 197 to all redox potentials to report data relative to the standard hydrogen electrode
  filter(avg_values != 'NAN') %>% 
  mutate(avg_values = as.numeric(avg_values)) %>% 
  dplyr::mutate(avg_values_fixed = avg_values + 197) %>% 
  mutate(datalogger = "4")%>% 
  separate(redox_avg, sep = "_Av", into = c("redox_depth", "avg_delete")) %>% 
  dplyr::select(-avg_delete)

easthydric_avg = 
  avg_separate(easthydric_dlname) %>% 
  left_join(easthydric_metadata, by = 'redox_NUM_Avg') %>% 
  mutate(redox_5_1.1_fromtip_Avg = as.character(redox_5_1.1_fromtip_Avg),
         redox_5_1.2_fromtip_Avg = as.character(redox_5_1.2_fromtip_Avg),
         redox_5_1.3_fromtip_Avg = as.character(redox_5_1.3_fromtip_Avg),
         redox_5_1.4_fromtip_Avg = as.character(redox_5_1.4_fromtip_Avg),
         redox_5_1.5_fromtip_Avg = as.character(redox_5_1.5_fromtip_Avg),
         redox_5_1.6_fromtip_Avg = as.character(redox_5_1.6_fromtip_Avg),
         redox_5_1.7_fromtip_Avg = as.character(redox_5_1.7_fromtip_Avg),
         redox_5_1.8_fromtip_Avg = as.character(redox_5_1.8_fromtip_Avg),
         redox_1A_Avg = as.character(redox_1A_Avg)) %>%  
  pivot_longer(-c("redox_NUM_Avg", "X", "redox_NUM_Std", "TIMESTAMP", "RECORD", "site", "position", "Betterdate"),
               names_to = "redox_avg", values_to = "avg_values") %>% 
  #add 197 to all redox potentials to report data relative to the standard hydrogen electrode
  filter(avg_values != 'NAN') %>% 
  mutate(avg_values = as.numeric(avg_values)) %>%  
  dplyr::mutate(avg_values_fixed = avg_values + 197) %>% 
  mutate(datalogger = "5")%>% 
  separate(redox_avg, sep = "_Av", into = c("redox_depth", "avg_delete")) %>% 
  dplyr::select(-avg_delete)

eastmesic_avg = 
  avg_separate(eastmesic_dlname) %>% 
  left_join(eastmesic_metadata, by = 'redox_NUM_Avg') %>% 
  pivot_longer(-c("redox_NUM_Avg", "X", "redox_NUM_Std", "TIMESTAMP", "RECORD", "site", "position", "Betterdate"),
               names_to = "redox_avg", values_to = "avg_values") %>% 
  #add 197 to all redox potentials to report data relative to the standard hydrogen electrode
  filter(avg_values != 'NAN') %>% 
  mutate(avg_values = as.numeric(avg_values)) %>% 
  dplyr::mutate(avg_values_fixed = avg_values + 197) %>% 
  mutate(datalogger = "1")%>% 
  separate(redox_avg, sep = "_Av", into = c("redox_depth", "avg_delete")) %>% 
  dplyr::select(-avg_delete)

eastdry_avg = 
  avg_separate(eastdry_dlname) %>% 
  left_join(eastdry_metadata, by = 'redox_NUM_Avg') %>% 
  mutate(redox_3_1.1_fromtip_Avg = as.character(redox_3_1.1_fromtip_Avg),
         redox_3_1.2_fromtip_Avg = as.character(redox_3_1.2_fromtip_Avg),
         redox_3_1.3_fromtip_Avg = as.character(redox_3_1.3_fromtip_Avg),
         redox_3_1.4_fromtip_Avg = as.character(redox_3_1.4_fromtip_Avg),
         redox_3_3.1_fromtip_Avg = as.character(redox_3_3.1_fromtip_Avg),
         redox_3_3.2_fromtip_Avg = as.character(redox_3_3.2_fromtip_Avg),
         redox_3_3.3_fromtip_Avg = as.character(redox_3_3.3_fromtip_Avg),
         redox_3_3.4_fromtip_Avg = as.character(redox_3_3.4_fromtip_Avg)
         ) %>% 
  pivot_longer(-c("redox_NUM_Avg", "X", "redox_NUM_Std", "TIMESTAMP", "RECORD", "site", "position", "Betterdate"),
               names_to = "redox_avg", values_to = "avg_values") %>% 
  #add 197 to all redox potentials to report data relative to the standard hydrogen electrode
  filter(avg_values != 'NAN') %>% 
  mutate(avg_values = as.numeric(avg_values)) %>% 
  dplyr::mutate(avg_values_fixed = avg_values + 197) %>% 
  mutate(datalogger = "3")%>% 
  separate(redox_avg, sep = "_Av", into = c("redox_depth", "avg_delete")) %>% 
  dplyr::select(-avg_delete)

```


```{r}

westhydric_std = 
  std_separate(westhydric_dlname) %>% 
  left_join(westhydric_metadata, by = 'redox_NUM_Std') %>% 
  pivot_longer(-c("redox_NUM_Std", "TIMESTAMP", "RECORD", "site", "position", "Betterdate"),
               names_to = "redox_std", values_to = "std_values") %>% 
  filter(redox_std != c("X", "redox_NUM_Avg"))%>% 
  separate(redox_std, sep = "_St", into = c("redox_depth", "std_delete")) %>% 
  dplyr::select(-std_delete)

westmesic_std = 
  std_separate(westmesic_dlname) %>% 
  left_join(westmesic_metadata, by = 'redox_NUM_Std') %>% 
  pivot_longer(-c("redox_NUM_Std", "TIMESTAMP", "RECORD", "site", "position", "Betterdate"),
               names_to = "redox_std", values_to = "std_values") %>% 
  filter(redox_std != c("X", "redox_NUM_Avg"))%>% 
  separate(redox_std, sep = "_St", into = c("redox_depth", "std_delete")) %>% 
  dplyr::select(-std_delete)

westdry_std = 
  std_separate(westdry_dlname) %>% 
  left_join(westdry_metadata, by = 'redox_NUM_Std') %>% 
  dplyr::select(-c(redox_NUM_Avg, X)) %>% 
  pivot_longer(-c("redox_NUM_Std", "TIMESTAMP", "RECORD", "site", "position", "Betterdate"),
               names_to = "redox_std", values_to = "std_values") %>% 
  filter(redox_std != c("NAN"))%>% 
  separate(redox_std, sep = "_St", into = c("redox_depth", "std_delete")) %>% 
  dplyr::select(-std_delete)

easthydric_std = 
  std_separate(easthydric_dlname) %>% 
  left_join(easthydric_metadata, by = 'redox_NUM_Std') %>% 
 dplyr::select(-c(redox_NUM_Avg, X)) %>% 
  dplyr::mutate(redox_5_1.1_fromtip_Std = as.character(redox_5_1.1_fromtip_Std),
         redox_5_1.2_fromtip_Std = as.character(redox_5_1.2_fromtip_Std),
         redox_5_1.3_fromtip_Std = as.character(redox_5_1.3_fromtip_Std),
         redox_5_1.4_fromtip_Std = as.character(redox_5_1.4_fromtip_Std),
         redox_5_1.5_fromtip_Std = as.character(redox_5_1.5_fromtip_Std),
         redox_5_1.6_fromtip_Std = as.character(redox_5_1.6_fromtip_Std),
         redox_5_1.7_fromtip_Std = as.character(redox_5_1.7_fromtip_Std),
         redox_5_1.8_fromtip_Std = as.character(redox_5_1.8_fromtip_Std),
          redox_1A_Std = as.character(redox_1A_Std)) %>%  
pivot_longer(-c("redox_NUM_Std", "TIMESTAMP", "RECORD", "site", "position", "Betterdate"), names_to = "redox_std", values_to = "std_values") %>% 
    filter(redox_std != c("NAN"))%>% 
  separate(redox_std, sep = "_St", into = c("redox_depth", "std_delete")) %>% 
  dplyr::select(-std_delete)


eastmesic_std = 
  std_separate(eastmesic_dlname) %>% 
  left_join(eastmesic_metadata, by = 'redox_NUM_Std') %>% 
   dplyr::select(-c(redox_NUM_Avg, X)) %>% 
  pivot_longer(-c("redox_NUM_Std", "TIMESTAMP", "RECORD", "site", "position", "Betterdate"),
               names_to = "redox_std", values_to = "std_values") %>% 
  filter(redox_std != c("X", "redox_NUM_Avg"))%>% 
  separate(redox_std, sep = "_St", into = c("redox_depth", "std_delete")) %>% 
  dplyr::select(-std_delete)

eastdry_std = 
  std_separate(eastdry_dlname) %>% 
  left_join(eastdry_metadata, by = 'redox_NUM_Std') %>% 
   dplyr::select(-c(redox_NUM_Avg, X)) %>% 
    mutate(redox_3_1.1_fromtip_Std = as.character(redox_3_1.1_fromtip_Std),
         redox_3_1.2_fromtip_Std = as.character(redox_3_1.2_fromtip_Std),
         redox_3_1.3_fromtip_Std = as.character(redox_3_1.3_fromtip_Std),
         redox_3_1.4_fromtip_Std = as.character(redox_3_1.4_fromtip_Std),
         redox_3_3.1_fromtip_Std = as.character(redox_3_3.1_fromtip_Std),
         redox_3_3.2_fromtip_Std = as.character(redox_3_3.2_fromtip_Std),
         redox_3_3.3_fromtip_Std = as.character(redox_3_3.3_fromtip_Std),
         redox_3_3.4_fromtip_Std = as.character(redox_3_3.4_fromtip_Std)
         ) %>% 
  pivot_longer(-c("redox_NUM_Std", "TIMESTAMP", "RECORD", "site", "position", "Betterdate"),
               names_to = "redox_std", values_to = "std_values") %>% 
    filter(redox_std != c("NAN"))%>% 
  separate(redox_std, sep = "_St", into = c("redox_depth", "std_delete")) %>% 
  dplyr::select(-std_delete)



```


```{r}
westhydric_merge =
  westhydric_avg %>% 
  left_join(westhydric_std)%>% 
  dplyr::select(-c(redox_NUM_Std)) %>% 
    filter(avg_values_fixed < 1000) %>%
  filter(avg_values_fixed > -600)  %>% 
     filter(std_values < 5) %>% 
    dplyr::select(-c(std_values)) 

westmesic_merge =
  westmesic_avg %>% 
  left_join(westmesic_std)%>% 
    dplyr::select(-c(redox_NUM_Std)) %>% 
    filter(avg_values_fixed < 1000) %>%
  filter(avg_values_fixed > -600)  %>% 
     filter(std_values < 5) %>% 
    dplyr::select(-c(std_values)) 


westdry_merge =
  westdry_avg %>% 
  left_join(westdry_std)%>% 
  dplyr::select(-c(redox_NUM_Std)) %>% 
    filter(avg_values_fixed < 1000) %>%
  filter(avg_values_fixed > -600)  %>% 
     filter(std_values < 5) %>% 
    dplyr::select(-c(std_values)) 


easthydric_merge =
  easthydric_avg %>% 
  left_join(easthydric_std)%>% 
  dplyr::select(-c(redox_NUM_Std)) %>% 
    filter(avg_values_fixed < 1000) %>%
  filter(avg_values_fixed > -600)  %>% 
     filter(std_values < 5) %>% 
    dplyr::select(-c(std_values)) 


eastmesic_merge =
  eastmesic_avg %>% 
  left_join(eastmesic_std)%>% 
  dplyr::select(-c(redox_NUM_Std)) %>% 
    filter(avg_values_fixed < 1000) %>%
  filter(avg_values_fixed > -600)  %>% 
     filter(std_values < 5) %>% 
    dplyr::select(-c(std_values)) 


eastdry_merge =
  eastdry_avg %>% 
  left_join(eastdry_std)%>% 
  dplyr::select(-c(redox_NUM_Std)) %>% 
    filter(avg_values_fixed < 1000) %>%
  filter(avg_values_fixed > -600)  %>% 
     filter(std_values < 5) %>% 
    dplyr::select(-c(std_values)) 


```



```

sensor names/depths

```{r}
sensor_depths_prename = read.csv("raw/2022_Sensor_depths.csv") 

sensor_depths = 
  sensor_depths_prename %>% 
  dplyr::rename(position = Site,
                site = Area) %>% 
  mutate(site = recode(site, "West" = "west",
                       "East" = "east")) %>% 
  mutate(position = recode(position, "Hydric" = "hydric",
                           "Mesic" = "mesic",
                           "Dry" = "dry"),
         Sensor = recode(Sensor, "redox_5-2-3_fromtip_Avg" = "redox_5_2-3_fromtip_Avg")) %>% 
  dplyr::select(-c(X2021_depth_cm, Change_from_2021_cm)) %>% 
  rename("depth_cm" = 'X2022_depth_cm') %>% 
  mutate(Sensor = str_remove(Sensor, "Redox_")) %>% 
  mutate(Sensor = str_remove(Sensor, "redox_")) %>% 
  mutate(Sensor = str_remove(Sensor, "_fromtip_Avg")) %>% 
  separate(Sensor, sep = "_", into = c("datalogger", "probe_sensor")) %>% 
  separate(probe_sensor, sep = "-", into = c("probe", "sensor")) %>% 
  mutate(sensor = as.numeric(sensor)) %>% 
  mutate(depth_cm = str_replace(depth_cm, "n.a.", "NA")) %>%
  mutate(depth_cm = as.numeric(depth_cm)) %>% 
  na.omit() 
  
  
```

```{r}


westhydric_depths = 
   depths_function(westhydric_merge)%>% 
   na.omit()

 
 easthydric_depths = 
   depths_function(easthydric_merge) %>% 
   na.omit()
 
 westmesic_depths = 
   depths_function(westmesic_merge)%>% 
   na.omit()
 
 eastmesic_depths = 
   depths_function(eastmesic_merge)%>% 
   na.omit()
 
 westdry_depths = 
   depths_function(westdry_merge)%>% 
   na.omit()
 
 eastdry_depths = 
   depths_function(eastdry_merge) %>% 
   na.omit()

```

Lag/lead cleaning

```{r}
westhydric_depths_outliers = 
   datacleaning_function(westhydric_depths) 

easthydric_depths_outliers = 
   datacleaning_function(easthydric_depths) 

westmesic_depths_outliers = 
  datacleaning_function(westmesic_depths)  

eastmesic_depths_outliers = 
  datacleaning_function(eastmesic_depths)  

westdry_depths_outliers = 
  datacleaning_function(westdry_depths) 
  
eastdry_depths_outliers = 
  datacleaning_function(eastdry_depths) 
 
```


filter std dev 
```{r}
alldata_cleaned =
  westhydric_depths_outliers %>% 
  vctrs::vec_c(easthydric_depths_outliers, westmesic_depths_outliers,
               eastmesic_depths_outliers, westdry_depths_outliers,
               eastdry_depths_outliers)

alldata_notcleaned =
  westhydric_depths %>% 
  vctrs::vec_c(easthydric_depths, westmesic_depths,
               eastmesic_depths, westdry_depths,
               eastdry_depths)

```



Output
```{r}

write.csv(alldata_cleaned, "processed/2022all_combine_temporary.csv")



```


plot
```{r}

  alldata_cleaned %>%
  # mutate(position = factor(position, levels = c("dry", "mesic", "hydric"))) %>%
  # mutate(site = recode(site, "east" = "acidic tundra",
  #                      "west" = "non-acidic tundra")) %>% 
  # mutate(site = factor(site, levels = c("non-acidic tundra", "acidic tundra"))) %>% 
  #filter(probe == 3) %>%
  ggplot(aes(x = as_datetime(Betterdate), y = avg_values_fixed))+
  geom_point(aes(color = depth_cm), alpha = 0.5)+
  #geom_line(aes(group = depth_cm, color = depth_cm, orientation = "x"))+
  scale_color_gradientn(colors = natparks.pals(name = "KingsCanyon", 8))+
  scale_x_datetime(date_breaks = "2 weeks" , date_labels = "%Y-%m-%d")+
  #scale_x_discrete(breaks = seq(-1,31,2))+
  labs(y = "redox potential, mV",
       x = "Post-cleaning",
       color = "depth, cm")+
  facet_grid(position~site)+
  theme_er1()+
  theme(legend.position = "top", axis.text.x = element_text (size = 10, vjust = 0.5, angle = 45))

ggsave("output/artifact_westhydric_fig_2022_std5.png", plot = artifact_westhydric_fig_2022_std5, height = 10, width = 10)
```





```{r}

artifact_westhydric_fig_2022 =
  westhydric_depths_outliers %>%
  # mutate(position = factor(position, levels = c("dry", "mesic", "hydric"))) %>%
  # mutate(site = recode(site, "east" = "acidic tundra",
  #                      "west" = "non-acidic tundra")) %>% 
  # mutate(site = factor(site, levels = c("non-acidic tundra", "acidic tundra"))) %>% 
  filter(probe == 3) %>%
  ggplot(aes(x = as_datetime(Betterdate), y = avg_values_fixed))+
  geom_point(aes(color = depth_cm), alpha = 0.5)+
  #geom_line(aes(group = depth_cm, color = depth_cm, orientation = "x"))+
  scale_color_gradientn(colors = natparks.pals(name = "KingsCanyon", 8))+
  scale_x_datetime(date_breaks = "2 weeks" , date_labels = "%Y-%m-%d")+
  #scale_x_discrete(breaks = seq(-1,31,2))+
  labs(y = "redox potential, mV",
       x = "Pre-cleaning",
       color = "depth, cm")+
  #facet_grid(position~site)+
  theme_er1()+
  theme(legend.position = "top", axis.text.x = element_text (size = 10, vjust = 0.5, angle = 45))

ggsave("output/artifact_westhydric_fig_2022.png", plot = artifact_westhydric_fig_2021, height = 10, width = 10)
```

