---
title: "Model-Data Comparisons"
author: "Erin C Rooney"
date: "2023-03-15"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Welcome

This is a data explainer that pulls in the necessary data files as well
as associated cleaning and grouping functions. Functions can be edited
in the "code/functions_for_modeldatacomparisons_dataexplainer.R" script.

View the easy to read published Markdown file
[here](https://rpubs.com/ecrooney/1019045).

### Known Issues

-   There are two types of information associated sample location: the
    moisture conditions and the tundra type. Sometimes the tundra type
    (non-acidic or acidic) is referred to as "area" and sometimes it's
    referred to as "site." Similarly, moisture conditions (dry, mesic,
    and hydric) are sometimes called "site" and sometimes called
    "position".

-   There are issues with capitalization. If dataframes aren't
    combining, it's a good chance that the issue is "Dry" and "dry"
    being unable to merge.

-   Sometimes tundra types are referred to as "west" and "east". West
    corresponds to non-acidic tundra and east corresponds to acidic
    tundra. This issue is the result of my data processing being focused
    on the figures/tables as output.

### Missing Data

I will add in the following datasets as we get them

-   Ferrous iron

-   Soil texture

-   rhizon ICP data for 2022

I need to add in the following datasets that we do have

-   pH of porewater

## Data Input and Processing

### Source Packages and Functions

```{r, warning=FALSE,message=FALSE,error=FALSE}
knitr::opts_chunk$set(echo = TRUE)

source("code/0-packages.R")
source("code/functions_for_modeldatacomparisons_dataexplainer.R")
```

### Sensor Data

#### Redox data

artifacts and frozen values removed and with hydrogen electrode
correction

```{r}
knitr::opts_chunk$set(echo = TRUE)

redox_withdepths2022 = read.csv("processed/allcombine_2022_frozen.csv")
redox_withdepths2021 = read.csv("processed/allcombine_2021_frozen.csv")

```

#### Moisture, Temperature, and Salinity

data with artifacts and frozen values removed and with hydrogen
electrode correction

```{r}
temp_sal_moist2021 = read.csv("processed/final_temp_salinity_avgs.csv")
temp_sal_moist2022 = read.csv("processed/2022final_temp_salinity_avgs.csv")
```

#### Soil properties

```{r}
thaw_depth2021 = read.csv("raw/thaw_depth_2021.csv")
thaw_depth2022 = read.csv("raw/thaw_depth_2022.csv")

bd_grav = read.csv("raw/PhIr2021_Soil_Inventory_bd.csv") 
```

### Porewater Data

All porewater data below is in a long format. I can include a function
for pivoting wider if that's helpful.

```{r}
#depths for sipper porewater samples
porewater_ICP_2022 = read.csv("processed/ICP_2022.csv")

#rhizon samplers (only one depth)
porewater_ICP_2021 = read.csv("processed/rhizon_2021.csv")


```

### Cleaning functions

#### Redox cleaning

Everything must be run sequentially

```{r}
#remove frozen data

#avg_values_fixed = redox potential (mV)

redox_unfrozen_2021 = redox2021_remove_frozenpoints_function(redox_withdepths2021)
  
redox_unfrozen_2022 = redox2022_remove_frozenpoints_function(redox_withdepths2022)
  
#group by season

#redox_avg_mV = redox potential (mV)
  
redox_grouped_2021 = redox_2021_groupbytemporal_function(redox_unfrozen_2021)

redox_grouped_2022 = redox_2022_groupbytemporal_function(redox_unfrozen_2022)

# figure test

redox_grouped_2021 %>% 
  mutate(month = factor(month, levels = c("early summer", "mid summer", "late summer", "early fall")))   %>%
  mutate(site = factor(site, levels = c("non-acidic tundra", "acidic tundra"))) %>% 
  ggplot(aes(y = depth_cm, x = redox_avg_mV, color = month, fill = month), group = 'position')+
  geom_point(size = 3, alpha = 0.8, shape = c(21))+
  geom_line(orientation = "y", show.legend = FALSE, linetype = "longdash")+
  geom_errorbar(aes(xmin=redox_avg_mV-redox_sd, xmax=redox_avg_mV+redox_sd), show.legend = FALSE)+
  scale_color_manual(values = (pnw_palette('Sunset', 4)))+
  scale_fill_manual(values = (pnw_palette('Sunset', 4)))+
  ylim(60, 0)+
  labs(x = '2021
       redox potential (mV)',
       y = "",
       color = "", fill = "")+
  scale_x_continuous(position="top")+
  facet_grid(position~site, switch = "x")+
  theme_er1()+
  theme(legend.position = "right",
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        strip.placement = "outside",
        axis.text.y=element_blank(),  #remove y axis labels
        axis.ticks.y=element_blank(),
        axis.text.x = element_text(size = 7.5, hjust=0.8,vjust=0.2,angle = 90))  #remove y axis ticks
#remove y axis ticks

# figure test passed. Data is good to go.

```

#### Moisture, Temperature, Salinity Cleaning

```{r}

#clean up dates, note data as frozen if below -1 C
moisturetempsal2021_cleaned = moisturetempsal_2021_cleaningfunction(temp_sal_moist2021)
moisturetempsal2022_cleaned = moisturetempsal_2022_cleaningfunction(temp_sal_moist2022)

#grouped temporal data

moisturetempsal2021_grouped = moisturetempsal_2021_groupfunction(moisturetempsal2021_cleaned)
moisturetempsal2022_grouped = moisturetempsal_2022_groupfunction(moisturetempsal2022_cleaned)

 moisturetempsal2021_grouped %>%
  mutate(month = factor(month, levels = c("early summer", "mid summer", "late summer", "early fall")))   %>%
  mutate(site = factor(site, levels = c("non-acidic tundra", "acidic tundra"))) %>%
  ggplot(aes(y = depth_cm, x = moisture_avg, color = month, fill = month), group = 'month')+
  geom_point(size = 4, alpha = 0.8, shape = c(21))+
  geom_line(orientation = "y", show.legend = FALSE, linetype = "longdash")+
  geom_errorbar(aes(xmin=moisture_avg-moisture_sd, xmax=moisture_avg+moisture_sd), show.legend = FALSE)+
  scale_color_manual(values = (pnw_palette('Sunset', 4)))+
  scale_fill_manual(values = (pnw_palette('Sunset', 4)))+
  ylim(60, 0)+
  labs(x = '2021
       soil moisture (%)',
       y = "depth (cm)",
       color = "", fill = "")+
  scale_x_continuous(position="top")+
  facet_grid(position~site, switch = "x")+
  theme_er1()+
  theme(legend.position = "none",
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        strip.placement = "outside",
        strip.text.y = element_blank(),
        axis.text.x = element_text(size = 7.5, hjust=0.8,vjust=0.2,angle = 90)) 

```

#### Soil properties cleaning

```{r}

#clean up data, rename areas for consistency, calculate volumetric moisture, simplify soil material type
bd_cleaned = soil_properties_cleaningfunction(bd_grav)

#select only necessary columns, simplify date-plot labels, set levels for material/horizons and date-plots, group by Area, Site, Horizon and summarise mean and sd for bulk density and volumetric water content

bd_grouped = soil_properties_groupingfunction(bd_cleaned)


#thaw depth cleaning

thawdepths_2021_cleaned = thawdepths_2021_cleaningfunction(thaw_depth2021)
thawdepths_2022_cleaned = thawdepths_2022_cleaningfunction(thaw_depth2022)


#thaw depth figure test

thawdepths_2021_cleaned %>% 
  mutate(Site = factor(Site, levels = c("Dry", "Mesic", "Hydric"))) %>% 
  #filter(Area == "non-acidic tundra") %>% 
  ggplot()+
  geom_violin(aes(x = as.Date(date2), y = thaw_depth_cm, group = as.Date(date2), fill = Site), alpha = 0.4)+
  labs(x = " ",
       y = "Thaw Depth, cm")+
  #geom_rect(aes(xmin=as_date('2021-06-15'), xmax= as_date('2021-08-09'), ymin=49.5, ymax=50.5), fill = "black")+
  #scale_fill_gradientn(colors = natparks.pals(name = "Banff"))+
  scale_fill_manual(values = c("#9a031e", "#40916c", "#118ab2"))+
  scale_x_date(date_breaks = "1 week", date_labels = "%b-%d-%Y")+
  ylim(90, 0)+
  facet_grid(Site~Area)+
  theme_er1()+
  theme(legend.position = "none", axis.text.x = element_text (vjust = 0.5, hjust=1, angle = 90, size = 9))

#bulk density figure test




```

#### Porewater cleaning

```{r}

#cleaning
porewater_ICP_2021_cleaned = rhizon_cleaningfunction(porewater_ICP_2021)

#2022 ICP sipper porewater data comes pre-cleaned

#grouping
porewater_ICP_2021_grouped = rhizon_groupingfunction(porewater_ICP_2021_cleaned)
porewater_ICP_2022_grouped = sipper_groupingfunction(porewater_ICP_2022)

#rhizon 2021 test figure

 porewater_ICP_2021_grouped %>%
  mutate(Area = recode(Area, "West" = "non-acidic tundra",
                       "East" = "acidic tundra")) %>% 
  mutate(Area = factor(Area, levels = c("non-acidic tundra", "acidic tundra"))) %>% 
  filter(ICP %in% c("Ca_ug/mL")) %>% 
  # mutate(ICP = recode(ICP, "calcium" = "calcium μg/mL",
  #                     "iron" = "iron μg/mL",
  #                     "aluminum" = "aluminum μg/mL",
  #                     "phosphorus" = "phosphorus μg/mL")) %>%   
  ggplot(aes(x = Betterdate, y = mean, color = Site)) +
  geom_line(aes(group = area_site, linetype = Area), size = 0.7, orientation = "x")+
  geom_point(aes(group = area_site, shape = Area), size = 3.5)+
  # scale_fill_gradientn(colors = (pnw_palette("Shuksan2")))+
  # scale_color_gradientn(colors = (pnw_palette("Shuksan2")))+
  scale_fill_manual(values = c("#f07167", "#a7c957", "#1e96fc", "#f07167", "#a7c957", "#1e96fc"))+
  scale_color_manual(values = c("#f07167", "#a7c957", "#1e96fc", "#f07167", "#a7c957", "#1e96fc"))+
  labs(color = "Moisture",
       linetype = "Acidity",
       y = "Concentration μg/mL",
       x = " ")+
  facet_grid(. ~ ICP, scales = "free_y") +
  theme_er1()+
  theme(axis.text.x = element_text(size = 8, hjust=0.25,vjust=0.2,angle = 45), legend.position = "none",
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank())

# sipper 2022 test figure
 
   porewater_ICP_2022_grouped %>% 
    filter(ICP == "Fe_ug_mL") %>% 
    mutate(Area = factor(Area, levels = c("non-acidic tundra", "acidic tundra"))) %>%
    mutate(Site = factor(Site, levels = c("Dry", "Mesic", "Hydric"))) %>% 
    ggplot(aes(x = mean, y = Depth_cm, fill = date_plot, group = date_plot)) +
    geom_errorbar(aes(xmin=(mean - sd), xmax=(mean + sd), color = date_plot))+
    geom_point(size = 3.5, shape = c(21), alpha = 0.6)+
    geom_line(aes(color = date_plot), orientation = "y", linetype = "longdash")+
    # scale_fill_gradientn(colors = (pnw_palette("Shuksan2")))+
    # scale_color_gradientn(colors = (pnw_palette("Shuksan2")))+
    scale_fill_manual(values = (pnw_palette("Sunset2", 15)))+
    scale_color_manual(values = (pnw_palette("Sunset2", 15)))+
    labs(fill = "Date",
         color = "Date",
         y = "",
         x = "Iron ug/mL")+
    scale_y_reverse()+
    scale_x_continuous(position = 'top') +
    facet_grid(Site ~ Area, scales = "free_x")+
    theme_er1()+
    theme(legend.position = "none",
          panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
          panel.background = element_blank(),
                strip.placement = "outside")

```

## Data Output

I usually analyze the data in the same script where I'm doing data
processing and cleaning, so these grouped and cleaned datasets do not
always have corresponding output files (other than the figures and
tables). I think the best approach is to create a modeling data folder
with output files from this data explainer (see below code chunk).

It will take some communication (I'll have to let you know if there are
any changes to processing code or data files so that you can re-run this
markdown file). But this file is helpful to me to because it simplifies
the processing that I'm doing which will be helpful for visitors to my
repository. So I have a vested interest in keeping it up to date.
Another option is for me to figure out how to combine markdown file with
a Drake outline, but I'm not at that stage yet.

```{r}

#Redox

write.csv(redox_withdepths2021, "modeling-data-output-files/redox_withdepths2021.csv")
write.csv(redox_withdepths2022, "modeling-data-output-files/redox_withdepths2022.csv")

write.csv(redox_unfrozen_2021, "modeling-data-output-files/redox_unfrozen_2021.csv")
write.csv(redox_unfrozen_2022, "modeling-data-output-files/redox_unfrozen_2022.csv")

write.csv(redox_grouped_2021, "modeling-data-output-files/redox_grouped_2021.csv")
write.csv(redox_grouped_2022, "modeling-data-output-files/redox_grouped_2022.csv")



#Moisture Temperature Salinity
write.csv(temp_sal_moist2021, "modeling-data-output-files/temp_sal_moist2021.csv")
write.csv(temp_sal_moist2022, "modeling-data-output-files/temp_sal_moist2022.csv")

write.csv(moisturetempsal2021_cleaned, "modeling-data-output-files/moisturetempsal2021_cleaned.csv")
write.csv(moisturetempsal2022_cleaned, "modeling-data-output-files/moisturetempsal2022_cleaned.csv")

write.csv(moisturetempsal2021_grouped, "modeling-data-output-files/moisturetempsal2021_grouped.csv")
write.csv(moisturetempsal2022_grouped, "modeling-data-output-files/moisturetempsal2022_grouped.csv")




#Porewater data
write.csv(porewater_ICP_2021, "modeling-data-output-files/porewater_ICP_2021.csv")
write.csv(porewater_ICP_2022, "modeling-data-output-files/porewater_ICP_2022.csv") #comes cleaned
write.csv(porewater_ICP_2021_cleaned, "modeling-data-output-files/porewater_ICP_2021_cleaned.csv")
write.csv(porewater_ICP_2021_grouped, "modeling-data-output-files/porewater_ICP_2021_grouped.csv")
write.csv(porewater_ICP_2022_grouped, "modeling-data-output-files/porewater_ICP_2022_grouped.csv")


#Soil Properties

write.csv(thawdepths_2021_cleaned, "modeling-data-output-files/thawdepths_2021_cleaned.csv")
write.csv(thawdepths_2022_cleaned, "modeling-data-output-files/thawdepths_2022_cleaned.csv")
write.csv(bd_grav, "modeling-data-output-files/bd_grav.csv")
write.csv(bd_cleaned, "modeling-data-output-files/bd_cleaned.csv")
write.csv(bd_grouped, "modeling-data-output-files/bd_grouped.csv")

```
