---
title: "Model-Data Comparisons"
author: "Erin C Rooney"
date: "2023-03-15"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Welcome

This is a data explainer that pulls in the necessary data files as well
as associated cleaning and grouping functions. Functions can be edited
in the "code/functions_for_modeldatacomparisons_dataexplainer.R" script.

### Known Issues

-   There are two types of information associated sample location: the
    moisture conditions and the tundra type. Sometimes the tundra type
    (non-acidic or acidic) is referred to as "area" and sometimes it's
    referred to as "site." Similarly, moisture conditions (dry, mesic,
    and hydric) are sometimes called "site" and sometimes called
    "position".

-   There are issues with capitalization. If dataframes aren't
    combining, it's a good chance that the issue is "Dry" and "dry"
    being unable to merge.

-   Sometimes tundra types are referred to as "west" and "east". West
    corresponds to non-acidic tundra and east corresponds to acidic
    tundra. This issue is the result of my data processing being focused
    on the figures/tables as output.

### Missing Data

I will add in the following datasets as we get them

-   Ferrous iron

-   Soil texture

-   rhizon ICP data for 2022

I need to add in the following datasets that we do have

-   pH of porewater


## Data Input and Processing

### Source Packages and Functions

```{r, warning=FALSE,message=FALSE,error=FALSE}
knitr::opts_chunk$set(echo = TRUE)

source("code/0-packages.R")
source("code/functions_for_modeldatacomparisons_dataexplainer.R")
```

### Sensor Data

#### Redox data 

artifacts and frozen values removed and with hydrogen electrode correction

```{r}
knitr::opts_chunk$set(echo = TRUE)

redox_withdepths2022 = read.csv("processed/allcombine_2022_frozen.csv")
redox_withdepths2021 = read.csv("processed/allcombine_2021_frozen.csv")

```

#### Moisture, Temperature, and Salinity 

data with artifacts and frozen values removed and with hydrogen electrode correction

```{r}
temp_sal_moist2021 = read.csv("processed/final_temp_salinity_avgs.csv")
temp_sal_moist2022 = read.csv("processed/2022final_temp_salinity_avgs.csv")
```

#### Soil properties

```{r}
thaw_depths = read.csv("raw/thaw_depth_2021.csv")

bd_grav = read.csv("raw/PhIr2021_Soil_Inventory_bd.csv") 

bulkdensity_tableoutput = read.csv("output/bd_grouped.csv")
```

### Porewater Data

All porewater data below is in a long format. I can include a function
for pivoting wider if that's helpful.

```{r}

porewater_ICP_2022 = read.csv("processed/ICP_2022.csv")

porewater_ICP_2021 = read.csv("processed/rhizon_2021.csv")

```

### Cleaning functions

#### Redox cleaning

```{r}
#remove frozen data
redox_unfrozen_2021 = redox2021_remove_frozenpoints_function(redox_withdepths2021)
  
redox_unfrozen_2022 = redox2022_remove_frozenpoints_function(redox_withdepths2022)
  
#group by season
  
redox_grouped_2021 = redox_2021_groupbytemporal_function(redox_unfrozen_2021)

redox_grouped_2022 = redox_2022_groupbytemporal_function(redox_unfrozen_2022)

```

#### Moisture, Temperature, Salinity Cleaning

```{r}

#clean up dates, note data as frozen if below -1 C
moisturetempsal2021_cleaned = moisturetempsal_2021_cleaningfunction(temp_sal_moist2021)
moisturetempsal2022_cleaned = moisturetempsal_2022_cleaningfunction(temp_sal_moist2022)

#grouped temporal data

moisturetempsal2021_grouped = moisturetempsal_2021_groupfunction(moisturetempsal2021_cleaned)
moisturetempsal2022_grouped = moisturetempsal_2022_groupfunction(moisturetempsal2022_cleaned)

```

#### Soil properties cleaning

```{r}

#clean up data, rename areas for consistency, calculate volumetric moisture, simplify soil material type
bd_cleaned = soil_properties_cleaningfunction(bd_grav)

#select only necessary columns, simplify date-plot labels, set levels for material/horizons and date-plots
bd_select = soil_properties_selectfunction(bd_cleaned)

```

## Data Output

I usually analyze the data in the same script where I'm doing data
processing and cleaning, so these grouped and cleaned datasets do not
always have corresponding output files (other than the figures and
tables). I think the best approach is to create a modeling data folder
with output files from this data explainer (see below code chunk). 

It will take some communication (I'll have to let you know if there are any changes to
processing code or data files so that you can re-run this markdown
file). But this file is helpful to me to because it simplifies the
processing that I'm doing which will be helpful for visitors to my
repository. So I have a vested interest in keeping it up to date.
Another option is for me to figure out how to combine markdown file with
a Drake outline, but I'm not at that stage yet.


```{r}

#Redox

write.csv(redox_withdepths2021, "modeling-data-output-files/redox_withdepths2021.csv")
write.csv(redox_withdepths2022, "modeling-data-output-files/redox_withdepths2022.csv")

write.csv(redox_unfrozen_2021, "modeling-data-output-files/redox_unfrozen_2021.csv")
write.csv(redox_unfrozen_2022, "modeling-data-output-files/redox_unfrozen_2022.csv")

write.csv(redox_grouped_2021, "modeling-data-output-files/redox_grouped_2021.csv")
write.csv(redox_grouped_2022, "modeling-data-output-files/redox_grouped_2022.csv")



#Moisture Temperature Salinity
write.csv(temp_sal_moist2021, "modeling-data-output-files/temp_sal_moist2021.csv")
write.csv(temp_sal_moist2022, "modeling-data-output-files/temp_sal_moist2022.csv")

write.csv(moisturetempsal2021_cleaned, "modeling-data-output-files/moisturetempsal2021_cleaned.csv")
write.csv(moisturetempsal2022_cleaned, "modeling-data-output-files/moisturetempsal2022_cleaned.csv")

write.csv(moisturetempsal2021_grouped, "modeling-data-output-files/moisturetempsal2021_grouped.csv")
write.csv(moisturetempsal2022_grouped, "modeling-data-output-files/moisturetempsal2022_grouped.csv")




#Porewater data
write.csv(porewater_ICP_2021, "modeling-data-output-files/porewater_ICP_2021.csv")
write.csv(porewater_ICP_2022, "modeling-data-output-files/porewater_ICP_2022.csv")


#Soil Properties

write.csv(thaw_depths, "modeling-data-output-files/thaw_depths.csv")
write.csv(bd_grav, "modeling-data-output-files/bd_grav.csv")
write.csv(bulkdensity_tableoutput, "modeling-data-output-files/bulkdensity_tableoutput.csv")

```
